A Bioconductor workflow for processing and analysing spatial proteomics data

### 2018-04-11 bioc-workflow_march2017.Rnw
all reviewers comments addressed

This procedure is also applicable to any packages, from
\href{https://cran.r-project.org/}{CRAN} as well as GitHub. Once a
package has been installed, it needs to be loaded for its
functionality to become available in the R session; this is done with
the \texttt{library} function e.g.  to load the \Biocpkg{pRoloc} one
would type \texttt{library("pRoloc")} after installation.

@@

While we focus here on a LOPIT-type dataset, these analyses are
\hl{relevent} for any quantitative spatial proteomics data, irrespective of
the fractionation or quantitation (i.e. labelled or label-free)
methods.

@@

Feature metadata typically contains general annotation about the
proteins (accession numbers, description, \ldots), information related
to the identification search (confidence scores, number of peptides,
\ldots) as well as annotation about know sub-cellular location (see in
particular the \textit{Markers} section) and results from data
analysis. The sample metadata would, for example, record what stable
isotope labels were used for the respective fraction (when labelled
quantitation is used), replicate number, fraction number along the
gradient and pooling information.

@@

To use the \texttt{readMSnSet2} function, as a minimum one must
specify the file path to the data and which columns of the spreadsheet
contain quantitation data. In the code chunk below, we start by
identifying the file that we want to use. The \texttt{system.file}
function is used to return the path to the \texttt{extdata} directory
from the \Biocexptpkg{pRolocdata} package, which is where our file of
interest resides. We then use the \texttt{dir} function to list the
content of that directory and store the path that matches the file
name of interest in the \texttt{csvfile}. Note that these two lines
are only needed here to locate a file in a package; in a more general
use case, the user would defined the \texttt{csvfile} variable
containing the file name of interest directly.

@@

<<combine, cache = cache>>=
combined <- combine(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)
combined
@

@@

\hl{This is further highlighted if we plot the
profiles of these two clusters using the \texttt{plotDist} function}
(Figure \ref{fig:plotDist2}). \hl{The \texttt{plotDist} function is
another useful visualisation that relies on marker annotation. It
allows one to represent the protein profiles occupancy along the
gradient. While the PCA
plot enables efficient visualisation of the complete dataset and
assessment the relative separation of different sub-cellular niches,
comparing profiles of a few marker clusters is useful to assess how
exactly they differ (in terms of peak channels, for example). On
figure} \ref{fig:plotDist2}, \hl{we plot the profile of the mitochondrial
and peroxisome markers to highlight the differences in profiles
between these two sets of markers in the channels labelled with tag
129C, as represented above along the 7th PC on the PCA plot on
figure} \ref{fig:plotmarkers}.

@@

The data can also be visualised along three PCs using the
\texttt{plot3D} function (Figure \ref{fig:plotmarkers3d}). When
produced interactively, the plot can be rotated and zoomined using the
mouse.

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object \hl{and then use the \textit{highlightOnPlot} function to 
visualise these}. For example, if we wanted to highlight \st{a} 
proteins with the accession numbers Q8CG48, Q8CG47, Q8K2Z4, and Q8C156, 
which are some of the proteins known to form part of the 13S condensin complex, 
we \hl{would call the following code:}

@@

Direct comparisons of individual channels in replicated experiments
do\st{es} not provide an adequate, goal-driven assessment of different
experiments. Indeed, due to the nature of the experiment and gradient
fraction collection, the quantitative channels do not correspond to
identical selected fractions along the gradient. For example, in the
table below (taken from \texttt{hl}'s \texttt{pData}) TMT channels
127C (among others) in both replicates originate from different sets
of gradient fractions (gradient fractions 7 - 9 and 8 - 9 for each
replicate, respectively). Different sets of gradient fractions are
often pooled to obtain enough material and optimise acurate
quantitation.

@@

\hl{The above analysis is comptationally intensive and best parallelised over
multiple workers. This phenoDisco analysis took ~24 hours to complete
when parallelised over 40 workers. As such, to avoid executing this
time consuming analysis users can access the above results which are
pre-computed and stored along with the \texttt{pRolocdata} package. 
Please see the appendix to load these results.}

@@

@@

We can plot the results using the \texttt{plot2D} function (Figure
\ref{fig:plotPDres}). The five new phenotype data clusters can be
extracted and examined.  In the code chunk below we write the results
to a .csv file using the \texttt{write.exprs} function. We use the
argument \texttt{fDataCols} to specify which columns of the
\texttt{featureData} to write.

@@

<<loadpdRes, echo=TRUE>>=
svmf <- dir(extdatadir, full.names = TRUE,
              pattern = "bpw-svmopt.rds") 
params <- readRDS(svmf)
@

@@

<<classify>>=
hl <- svmClassification(hl, params, class.weights = w, fcol = "markers")
@

@@

<<classify2, eval = FALSE>>=
hl <- svmClassification(hl, cost = 16, sigma = 0.1, class.weights = w, fcol = "markers")
@

@@

In the example below we \hl{first} remove these 4 classes from the marker set
\hl{and create a new marker set called tlmarkers. Then we}
re-run the \texttt{knnOptimisation} for each data source and then run
the \texttt{knntlOptimisation} with the 10 remaining classes. (Note:
this is not run live as the \texttt{hl} dataset with 10 classes,
707 markers and 4$^{10}$ combinations of parameters takes around 76
hours to run on the University of Cambridge HPC using 256 workers. 
\hl{To save time for users, the results of the following optimisation 
are pre-computed and provided with the \texttt{pRolocdata} package. 
Please see the appendix for details on how
to load these directly}.)

\st{To remove the 4 classes and create a new column of markers in the
feature data called tlmarkers to use for the analysis:}

@@

We generally find supervised learning more suited to the task of
protein localisation prediction in which we use high-quality curated
marker proteins to build a classifier, instead of using an entirely
unsupervised approach to look for clusters and then look for
enrichment of organelles and complexes. In the latter we do not make
good use of valuable prior knowledge, and in our experience
unsupervised clustering can be extremely difficult due to poor
estimates of the number of clusters that may appear in the data.

@@

\subsubsection{Novelty detection results}

@@

\subsubsection{SVM optimisation}

@@

\subsubsection{Transfer learning results}

>>>>

This procedure is also applicable to any packages, from
\href{https://cran.r-project.org/}{CRAN} as well as GitHub. Once a
package has been installed, it needs to be loaded for its
functionality to become available in the R session; this is done with
the \texttt{library} function e.g.  to load the \Biocpkg{pRoloc} package one
would type \texttt{library("pRoloc")} after installation.

@@

While we focus here on a LOPIT-type dataset, these analyses are
\hl{relevent} for any quantitative spatial proteomics data,  
irrespective of the fractionation (i.e. density gradient or 
differential centrifugation \cite{Itzhak:2016}) or quantitation 
(i.e. labelled or label-free) methods.

@@

Feature metadata typically contains general annotation about the
proteins (accession numbers, description, \ldots), information related
to the identification search (confidence scores, number of peptides,
\ldots) as well as annotation about known sub-cellular location (see in
particular the \textit{Markers} section) and results from data
analysis. The sample metadata would, for example, record what stable
isotope labels were used for the respective fraction (when labelled
quantitation is used), replicate number, fraction number along the
gradient and pooling information.

@@

To use the \texttt{readMSnSet2} function, as a minimum one must
specify the file path to the data and which columns of the spreadsheet
contain quantitation data. In the code chunk below, we start by
identifying the file that we want to use. The \texttt{system.file}
function is used to return the path to the \texttt{extdata} directory
from the \Biocexptpkg{pRolocdata} package, which is where our file of
interest resides. We then use the \texttt{dir} function to list the
content of that directory and store the path that matches the file
name of interest in the \texttt{csvfile}. Note that these two lines
are only needed here to locate a file in a package; in a more general
use case, the user would define the \texttt{csvfile} variable
containing the file name of interest directly.

@@

<<combine, cache = cache>>=
combined <- MSnbase::combine(hyperLOPIT2015ms3r1, hyperLOPIT2015ms3r2)
combined
@

@@

\hl{This is further highlighted if we plot the
profiles of these two clusters using the \texttt{plotDist} function}
(Figure \ref{fig:plotDist2}). \hl{The \texttt{plotDist} function is
another useful visualisation that relies on marker annotation. It
allows one to represent the protein profiles occupancy along the
gradient. While the PCA
plot enables efficient visualisation of the complete dataset and
assessment the relative separation of different sub-cellular niches,
comparing profiles of a few marker clusters is useful to assess how
exactly they differ (in terms of peak channels, for example). On
figure} \ref{fig:plotDist2}, \hl{we plot the profiles of the 
mitochondrial and peroxisome markers to highlight the differences 
in the channels labelled with tag 129C, also represented above  
along the 7th PC on the PCA plot on figure} \ref{fig:plotmarkers}.

@@

The data can also be visualised along three PCs using the
\texttt{plot3D} function (Figure \ref{fig:plotmarkers3d}). When
produced interactively, the plot can be rotated and zoomed using the
mouse.

@@

\subsection*{Features of interest}
In addition to adding annotation using the \texttt{addMarkers}
function, one can store specific sets of proteins by using the
\textit{Features of interest} infrastructure from the
\Biocpkg{MSnbase} package. If users have specific subsets of proteins
they wish to highlight in their data (possibly across multiple
experiments) they would first create a \texttt{FeaturesOfInterest}
object \hl{and then use the \textit{highlightOnPlot} function to 
visualise these}. For example, if we wanted to highlight 
proteins with the accession numbers Q8CG48, Q8CG47, Q8K2Z4, and 
Q8C156, which are some of the proteins known to form part of the 
13S condensin complex, we \hl{would call the following code:}

@@

Direct comparisons of individual channels in replicated experiments
do not provide an adequate, goal-driven assessment of different
experiments. Indeed, due to the nature of the experiment and gradient
fraction collection, the quantitative channels do not correspond to
identical selected fractions along the gradient. For example, in the
table below (taken from \texttt{hl}'s \texttt{pData}) TMT channels
127C (among others) in both replicates originate from different sets
of gradient fractions (gradient fractions 7 - 9 and 8 - 9 for each
replicate, respectively). Different sets of gradient fractions are
often pooled to obtain enough material and optimise acurate
quantitation.

@@

\hl{The above analysis is computationally intensive and best parallelised over
multiple workers. This phenoDisco analysis took ~24 hours to complete
when parallelised over 40 workers. As such, in the interest of time 
users can access the above results which are
pre-computed and stored along with the \texttt{pRolocdata} package. 
Please see the appendix to load these results.}

@@

We can plot the results using the \texttt{plot2D} function (Figure
\ref{fig:plotPDres}).

@@

The five new phenotype data clusters can be
extracted and examined.  In the code chunk below we write the results
to a .csv file using the \texttt{write.exprs} function. We use the
argument \texttt{fDataCols} to specify which columns of the
\texttt{featureData} to write.

@@

<<loadpdRes, echo=FALSE>>=
svmf <- dir(extdatadir, full.names = TRUE,
              pattern = "bpw-svmopt.rds") 
params <- readRDS(svmf)
@

@@

<<classify, warning = FALSE>>=
hl <- svmClassification(hl, params, class.weights = w, fcol = "markers")
@

@@

<<classify2, eval = FALSE, warning = FALSE>>=
hl <- svmClassification(hl, cost = 16, sigma = 0.1, class.weights = w, fcol = "markers")
@

@@

In the example below we \hl{first} remove these 4 classes from the marker set
\hl{and create a new marker set called tlmarkers. Then we}
re-run the \texttt{knnOptimisation} for each data source and then run
the \texttt{knntlOptimisation} with the 10 remaining classes. (Note:
this is not run live as the \texttt{hl} dataset with 10 classes,
707 markers and 4$^{10}$ combinations of parameters takes around 76
hours to run on the University of Cambridge HPC using 256 workers. 
\hl{To save time for users, the results of the following optimisation 
are pre-computed and provided with the \texttt{pRolocdata} package. 
Please see the appendix for details on how
to load these directly}.)q

@@

We generally find supervised learning more suited to the task of
protein localisation prediction in which we use high-quality curated
marker proteins to build a classifier, instead of using an entirely
unsupervised approach to look for clusters and then look for
enrichment of organelles and complexes. In the latter we do not make
good use of valuable prior knowledge, and in our experience
unsupervised clustering can be extremely difficult due to \hl{
(i) the loose definition of what constitutes a cluster (for example 
whether it is defined by the quantitative data or the localisation 
information), (ii) the influence of the algorithm assumption on the 
cluster identification (for example parametric or non-parametric) 
and (iii) poor estimates of the number of clusters that may appear 
in the data.}

@@

\subsubsection*{Novelty detection results}

@@

\subsubsection*{SVM optimisation}

@@

\subsubsection*{Transfer learning results}
